---
title: "Create ZCTA Crossrefs"
author: "Tyler Schappe"
date: "2022-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(tigris_use_cache = TRUE)
```

```{r}
library(tidyverse)
library(tigris)
library(sf)
```

# Merging ZCTAs with CTs

**Notes:**

  - We will be given a patient ZIP
  - This can correspond to a ZCTA, which is a collection of census blocks
  - We want to use a deprivation index at the census tract level
  - The problem is there are multiple CTs per ZCTA
  - US Census has a cross-ref file but it is one ZCTA to all overlapping CTs, with no indication for how much they overlap

**Approach:** Our approach is to loop over census tracts and find the ZCTA for which it spatially overlaps most. 

## Test on Just 2 ZIPs

### ZCTA Polygons

Get ZCTA polygons
```{r}
zcta.cary <- tigris::zctas(starts_with = "2751")
zcta.cary <- zcta.cary[zcta.cary$ZCTA5CE20 %in% c("27518", "27511") ,]
```

Project to US Albers and make valid
```{r}
zcta.cary <-
  zcta.cary %>%
  st_transform(crs = "EPSG:5070") %>%
  st_make_valid()
```

### Census Tracts

Get CT polygons in NC
```{r}
ct.nc <- tigris::tracts(state = "NC", year = 2020)
```

Project to US Albers and make valid
```{r}
ct.nc <- 
  ct.nc %>%
  st_transform(crs = "EPSG:5070") %>%
  st_make_valid()
```


### Join Without "Largest"

This is just to visualize a basic join.

```{r}
zcta.cary.ct <- st_join(zcta.cary, ct.nc)
```

#### Map

```{r}
ggplot(data = zcta.cary.ct)+
  #27518
  geom_sf(data = zcta.cary.ct[zcta.cary.ct$ZCTA5CE20 == "27518",],
          fill = "red",
          alpha = 0.01
  )+
  geom_sf(data = ct.nc[ct.nc$GEOID %in% zcta.cary.ct$GEOID[zcta.cary.ct$ZCTA5CE20 == "27518"] ,],
          fill = "blue",
          alpha = 0.1
  )+
  #27511
  geom_sf(data = zcta.cary.ct[zcta.cary.ct$ZCTA5CE20 == "27511",],
          fill = "yellow",
          alpha = 0.01
  )+
  geom_sf(data = ct.nc[ct.nc$GEOID %in% zcta.cary.ct$GEOID[zcta.cary.ct$ZCTA5CE20 == "27511"] ,],
          fill = "red",
          alpha = 0.1
  )
```

```{r}
sum(duplicated(zcta.cary.ct$GEOID))
```

**Result:** CTs that overlap both ZCTAs were assigned to both

### Join With "Largest"

**Note:** We have the ZCTAs listed first in the join.

```{r}
zcta.cary.ct.largest <- st_join(zcta.cary, ct.nc, largest = TRUE)
```

#### Map

```{r}
ggplot(data = zcta.cary.ct.largest)+
  #27518
  geom_sf(data = zcta.cary.ct.largest[zcta.cary.ct.largest$ZCTA5CE20 == "27518",],
          fill = "red",
          alpha = 0.01
  )+
  geom_sf(data = ct.nc[ct.nc$GEOID %in% zcta.cary.ct.largest$GEOID[zcta.cary.ct.largest$ZCTA5CE20 == "27518"] ,],
          fill = "blue",
          alpha = 0.1
  )+
  #27511
  geom_sf(data = zcta.cary.ct.largest[zcta.cary.ct.largest$ZCTA5CE20 == "27511",],
          fill = "yellow",
          alpha = 0.01
  )+
  geom_sf(data = ct.nc[ct.nc$GEOID %in% zcta.cary.ct.largest$GEOID[zcta.cary.ct.largest$ZCTA5CE20 == "27511"] ,],
          fill = "red",
          alpha = 0.1
  )
```

**Result:** Since we have the ZCTAs listed first, "largest" finds the CT that overlaps each ZCTA most (ie. only 2 CTs). This is not what we want -- we need to switch the order.



### Inverse Join With "Largest"

**Note:** We put the CTs first in the join so that "largest" finds the ZCTA that overlaps each CT most.

```{r}
ct.nc.cary.zcta <- st_join(ct.nc, zcta.cary, largest = TRUE)

ct.nc.cary.zcta <- ct.nc.cary.zcta[!is.na(ct.nc.cary.zcta$ZCTA5CE20) ,]
```

#### Map

```{r}
ggplot(data = ct.nc.cary.zcta)+
  #27518
  geom_sf(data = zcta.cary[zcta.cary$ZCTA5CE20 == "27518",],
          fill = "red",
          alpha = 0.1
  )+
  geom_sf(data = ct.nc.cary.zcta[ct.nc.cary.zcta$GEOID %in% ct.nc.cary.zcta$GEOID[ct.nc.cary.zcta$ZCTA5CE20 == "27518"] ,],
          fill = "blue",
          alpha = 0.1
  )+
  #27511
  geom_sf(data = zcta.cary[zcta.cary$ZCTA5CE20 == "27511",],
          fill = "green",
          alpha = 0.1
  )+
  geom_sf(data = ct.nc.cary.zcta[ct.nc.cary.zcta$GEOID %in% ct.nc.cary.zcta$GEOID[ct.nc.cary.zcta$ZCTA5CE20 == "27511"] ,],
          fill = "yellow",
          alpha = 0.1
  )+
  theme_bw()
```

Look for CT GEOIDs
```{r}
sum(duplicated(ct.nc.cary.zcta$GEOID))
```

**Result:** It works -- use this.

```{r}
ggplot(data = ct.nc.cary.zcta)+
  #27518
  # geom_sf(data = zcta.cary[zcta.cary$ZCTA5CE20 == "27518",],
  #         fill = "red",
  #         alpha = 0.1
  # )+
  # geom_sf(data = ct.nc.cary.zcta[ct.nc.cary.zcta$GEOID %in% ct.nc.cary.zcta$GEOID[ct.nc.cary.zcta$ZCTA5CE20 == "27518"] ,],
  #         fill = "blue",
  #         alpha = 0.1
  # )+
  #27511
  geom_sf(data = zcta.cary[zcta.cary$ZCTA5CE20 == "27511",],
          fill = "green",
          alpha = 0.1
  )+
  geom_sf(data = ct.nc.cary.zcta[ct.nc.cary.zcta$GEOID %in% ct.nc.cary.zcta$GEOID[ct.nc.cary.zcta$ZCTA5CE20 == "27511"] ,],
          fill = "yellow",
          alpha = 0.1
  )
```

